SparseIDS: Learning Packet Sampling with Reinforcement Learning

### 2020-05-28 paper_rl/presentation.tex
Incorporate Joachim's remarks

\author[Maximilian Bachl]{%
	Maximilian Bachl\email{maximilian.bachl@tuwien.ac.at} \and Fares Meghdouri\email{fares.meghdouri@tuwien.ac.at} \and Tanja Zseby\email{tanja.zseby@tuwien.ac.at} \and Joachim Fabini\email{joachim.fabini@tuwien.ac.at}
}

@@

\begin{frame}{Recurrent Neural Networks (RNNs)}
\begin{itemize}
\item RNNs are neural networks for \textbf{sequences}
\item Good performance for network traffic (\textit{Explainability and Adversarial Robustness for RNNs}, Hartl, Bachl, Fabini, Zseby, 2020) with \textbf{supervised} training
\item Can detect intrusion \textbf{before} attack is over
\end{itemize}
\end{frame}

\begin{frame}{Sampling for RNNs}
\begin{itemize}
\item Processing every packet by a neural network is \textbf{resource-intensive}
\item Often, attack is apparent after the first packets of a flow $\rightarrow$ not \textbf{necessary} to continue inspecting packets
\item Some packets might contain \textbf{no information} that is useful for classification
\end{itemize}
\pause
\begin{block}{Insight:}
Only process packets if the are useful for the classifier! 
\end{block}
\end{frame}

@@

\begin{frame}{Objective function}
Optimize
\begin{align*}
\text{Objective} = \text{Accuracy} + \alpha \times \text{Sparsity}
\end{align*}
\end{frame}

\begin{frame}{Neural Network architecture}
Implement using \textbf{Reinforcement Learning} based on the \textbf{Actor Critic} approach:
\begin{itemize}
\item \textbf{Classifier:} Decides whether a flow is benign or an attack at each packet
\item \textbf{Critic Network:} \textbf{Estimates} the \textbf{accuracy} and \textbf{sparsity} for the current flow
\item \textbf{Actor network:} Outputs the number of \textbf{packets to skip} to maximize objective based on Critic's output
\end{itemize}
\end{frame}

@@

\begin{frame}{Continuous vs discrete actions}
\begin{itemize}
\item \textbf{Current techniques} for Natural Language Processing allow skipping up to \textbf{fixed number} of steps (like 20).
\begin{itemize}
\item Using a categorical probability distribution.
\end{itemize}
\item Our goal: \textbf{skip arbitrarily many} packets. 
\begin{itemize}
\item Need a continuous distribution that can't output $<0$ because e.g.~skipping -1 packet is impossible.
\end{itemize}
\end{itemize}
\pause
\begin{block}{Insight:}
Use log-normal probability distribution! 
\end{block}
\end{frame}

@@

We implement:
\begin{itemize}
\item \textbf{random}: take packets with equal probability
\item \textbf{every \textit{i}th}: take every $i$th packet
\item \textbf{relative first \textit{m}}: for example: take the first 20\% of packets of each flow
\item \textbf{first \textit{m}}: for example: take the first 10 packets of each flow
\end{itemize}
\end{frame}

\begin{frame}{Comparing with other sampling methods}
\centering
\begin{tabular*}{\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
& RL & random & every $i$th & relative first \emph{m} & first \emph{m} \\	\midrule
Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
Precision & \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
\bottomrule
\end{tabular*}
\end{frame}

@@

\begin{frame}{Steering}
Set a maximum percentage of packets to be sampled (like 50\%). Then
\begin{itemize}
\item \textbf{initialize} RL with \textbf{high tradeoff} $\alpha$
\item \textbf{continuously decrease} $\alpha$ towards 0
\item \textbf{stop} if maximum percentage reached
\end{itemize}
\end{frame}

@@

\section{Discussion}
\begin{frame}{Conclusions}
\begin{itemize}
\item Reinforcement Learning \textbf{works} very well for sampling
\item \textbf{Sharing} all neural network weights makes our approach \textbf{very lightweight}
\item \textbf{Continuous actions} enable skipping \textbf{arbitrarily many} packets and work well
\item \textbf{Steering} allows using a \textbf{fixed cost} budget and then guide the RL to achieve \textbf{maximum accuracy} for that budget
\item \textbf{Other applications} for traffic classification and other networking tasks?
\end{itemize}
\end{frame}

>>>>

\author[M. Bachl et al.]{%
	\underline{Maximilian Bachl}\email{maximilian.bachl@tuwien.ac.at} \and Fares Meghdouri \and Tanja Zseby \and Joachim Fabini
}

@@

\begin{frame}{Recurrent Neural Networks (RNNs)}
\begin{itemize}
\item RNNs are neural networks for \textbf{sequences}
\item Good performance for network traffic \footnote{\textit{Explainability and Adversarial Robustness for RNNs}, Hartl, Bachl, Fabini, Zseby, 2020} with \textbf{supervised} training
\item Can detect intrusion \textbf{before} an attack is over
\end{itemize}
\end{frame}

\begin{frame}{Sampling for RNNs}
\begin{itemize}
\item Processing every packet by a neural network is \textbf{resource-intensive}
\item Often, attack is apparent after the first packets of a flow $\rightarrow$ not \textbf{necessary} to continue inspecting packets
\item Some packets might contain \textbf{no information} that is useful for classification
\end{itemize}
\pause
\begin{block}{Goal:}
Only process packets if they are useful for the classifier! 
\end{block}
\end{frame}

@@

\begin{frame}{Objective function}
Optimize
\begin{align*}
\text{Objective} = \text{Accuracy} + \alpha \times \text{Sparsity}
\end{align*}
where Sparsity is $\frac{\text{sampled packets}}{\text{all packets}}$ and $\alpha$ is a configurable tradeoff parameter. 
\end{frame}

\begin{frame}{Neural Network architecture}
Implement using \textbf{Reinforcement Learning} based on the \textbf{Actor Critic} approach:
\begin{itemize}
\item \textbf{Classifier:} Decides whether a flow is benign or an attack at each packet
\item \textbf{Critic Network:} \textbf{Estimates} the \textbf{accuracy} and \textbf{sparsity} for the current flow
\item \textbf{Actor network:} Outputs the number of subsequent \textbf{packets to skip} to maximize objective based on Critic's output
\end{itemize}
\end{frame}

@@

\begin{frame}{Continuous vs discrete actions}
\begin{itemize}
\item \textbf{Current techniques} for Natural Language Processing allow skipping up to \textbf{fixed number} of steps (like 20).
\begin{itemize}
\item Using a categorical probability distribution.
\end{itemize}
\item Our goal: \textbf{skip arbitrarily many} packets. 
\begin{itemize}
\item Need a continuous distribution that can't output $<0$ because e.g.~skipping -1 packet is not useful.
\end{itemize}
\end{itemize}
\pause
\begin{block}{Solution:}
Use log-normal probability distribution! 
\end{block}
\end{frame}

@@

Comparison against:
\begin{itemize}
\item \textbf{random}: take packets with equal probability
\item \textbf{every \textit{i}th}: take every $i$th packet
\item \textbf{relative first \textit{m}}: for example: take the first 20\% of packets of each flow (not possible if flow length isn't known in advance)
\item \textbf{first \textit{m}}: for example: take the first 10 packets of each flow
\end{itemize}
\end{frame}

\begin{frame}{Comparing with other sampling methods}
\centering
\begin{tabular*}{1.05\columnwidth}{>{\rowmac}l @{\extracolsep{\fill}} >{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c>{\rowmac}c<{\clearrow}} \toprule
& RL & random & every $i$th & rel.~first \emph{m} & first \emph{m} \\	\midrule
Sparsity & 76.3\% & 76.3\% & 76.3\% & 76.3\% & 76.3\% \\ \midrule
Accuracy & \textbf{99.4\%} & 96.9\% & 97.8\% & 97.3\% & 98.3\% \\
Precision& \textbf{98.5\%} & 92.4\% & 95.8\% & 93.8\% & 95.6\% \\
Recall & \textbf{99.0\%} & 95.6\% & 95.4\% & 95.5\% & 97.9\% \\
F1 & \textbf{99.7\%} & 94.0\% & 95.6\% & 94.7\% & 96.7\% \\
Youden & \textbf{98.5\%} & 93.0\% & 93.9\% & 93.4\% & 96.4\% \\
\bottomrule
\end{tabular*}
\end{frame}

@@

\begin{frame}{Steering}
Set a maximum percentage of packets to be sampled (like 50\%) as a \textbf{budget}. Then
\begin{itemize}
\item \textbf{initialize} RL with \textbf{high tradeoff} $\alpha$
\item \textbf{continuously decrease} $\alpha$ towards 0
\item \textbf{stop} if maximum percentage reached
\end{itemize}
\end{frame}

@@

\section{Discussion}
\begin{frame}{Conclusions}
\begin{itemize}
\item Reinforcement Learning \textbf{works} very well for sampling
\item \textbf{Sharing} all neural network weights makes our approach \textbf{very lightweight}
\item \textbf{Continuous actions} enable skipping \textbf{arbitrarily many} packets and work well
\item \textbf{Steering} allows using a \textbf{fixed cost} budget and then guide the RL to achieve \textbf{maximum accuracy} for that budget
\item \textbf{Other applications} of RL-based packet sampling for traffic classification and other networking tasks?
\end{itemize}
\end{frame}
